---
title: "power_prop.test"
output:
  pdf_document: default
  html_document: default
---
# --- Settings ---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# --- Packages ---

```{r}
library(tidyverse)

```

# --- Analysis ---

I want to compute achieved power, as a function of:
-- (1) sample size
-- (2) proportion 1
-- (3) proportion 2

I will only explore a small number, although large range, of sample sizes: 50, 100, 500, 1000 (per cell). I will explore all possible proportions, with a step size of 0.10.

# 1. create data frame to store results

```{r}
n_vals <- c(50, 100, 500, 1000)
prop1 <- seq(0, 1, 0.1)
prop2 <- seq(0, 1, 0.1)

total_rows <- length(n_vals) * length(prop1) * length(prop2)

df_power = data.frame(n = integer(total_rows),
                      p1 = double(total_rows),
                      p2 = double(total_rows),
                      power = double(total_rows),
                      power_bonf = double(total_rows))

df_power %>% 
  sample_n(10)
```

# 2. loop through all values and compute achieved power

```{r}
num_tests <- choose(n = 5, k = 2)

i <- 0
for (n_i in n_vals){
  for (p1_i in prop1){
    for (p2_i in prop2){
      i <- i + 1
      
      df_power[i,]$n <- n_i
      df_power[i,]$p1 <- p1_i
      df_power[i,]$p2 <- p2_i
      df_power[i,]$power <- power.prop.test(n = n_i,
                                            p1 = p1_i,
                                            p2 = p2_i,
                                            sig.level = 0.05,
                                            alternative = "two.sided")$power
      df_power[i,]$power_bonf <- power.prop.test(n = n_i,
                                                p1 = p1_i,
                                                p2 = p2_i,
                                                sig.level = 0.05/num_tests,
                                                alternative = "two.sided")$power
      
      
    }
  }
}

```

# 3. look at computed values

```{r}
df_power %>% 
  sample_n(10)
```


# 4. graph results

The thing to observe here is basically that with a sample size of n=100 per cell. Power is extremely low, for difference between the two proportions that are of size 10% (i.e. prop 2 is 10% percentage points than prop 1, like say 60% v. 50%). Power becomes reasonable (~80% and over), when the difference is of size 20% (e.g. if n=100, prop1 = 50, and prop2 = 70%, than power = 83%).

```{r, fig.width=12, fig.height=10}
plot_power_prop.test <-
df_power %>% 
  mutate(power = round(power, 2)) %>% 
  ggplot(aes(x = p1,
             y = p2,
             fill = power,
             label = power)) +
  geom_raster() +
  geom_text() +
  facet_wrap(vars(n), ncol = 2, scales = "free") +
  scale_x_continuous(breaks = prop1) +
  scale_y_continuous(breaks = prop2) +
  labs(x = "Proportion 1",
       y = "Proportion 2",
       title = paste0("Power as a function of:",
                      "\n (1) Proportion 1 (x-axis)",
                      "\n (2) Proportion 2 (y-axis)",
                      "\n (3) Sample Size, per cell (panels)")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

plot_power_prop.test

```

# 5. save graph

```{r}
if (FALSE){
ggsave(plot = plot_power_prop.test,
       filename = "plot_power_prop.test.pdf",
       width = 12, height = 10)
}
```

# 6. graph same results, but with alpha level adjusted for bonferroni correction

Now, here the differences between prop 1 and prop 2 in percentage points has to be of size 30% for there to be ~80%/90%+ power.

```{r, fig.width=12, fig.height=10}
df_power %>% 
  mutate(power_bonf = round(power_bonf, 2)) %>% 
  ggplot(aes(x = p1,
             y = p2,
             fill = power_bonf,
             label = power_bonf)) +
  geom_raster() +
  geom_text() +
  facet_wrap(vars(n), ncol = 2, scales = "free") +
  scale_x_continuous(breaks = prop1) +
  scale_y_continuous(breaks = prop2) +
  labs(x = "Proportion 1",
       y = "Proportion 2",
       title = paste0("Power as a function of:",
                      "\n (1) Proportion 1 (x-axis)",
                      "\n (2) Proportion 2 (y-axis)",
                      "\n (3) Sample Size, per cell (panels)",
                      "\n (note: this is power, after adjust p-level for Bonferroni correction")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))


```

