---
title: "prop_dependent"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# --- Settings ---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
plot_save <- FALSE
```


# --- Packages ---

```{r}
library(readxl)
library(tidyverse)
library(PropCIs)
library(gee)

```

# --- Load Data ---

# 1. read

```{r}
ex_raw <- read.table("http://www.stat.ufl.edu/~aa/cat/data/Envir_opinions.dat", header=TRUE)
```

# 2. view

```{r}
View(ex_raw)
ex_raw
```

# 3. shape

```{r}
# make tabular
ex <- xtabs(data = ex_raw,
            formula = ~y1 + y2)

# view
ex
```
# --- Analysis ---

# 1. mcnemar test

- test if there is dependency between two categorical variables (i.e. whether is evidence against marginal homogeneity).
- see, Agresti, Chapter 8, p. 227-228
- so: result, is that we don't have evidence against marginal homogeneity. which makes sense. props look close

```{r}
mcnemar.test(x = ex,
             correct = FALSE)
```

```{r}
addmargins(ex)

colSums(prop.table(ex)) # yes & no on taxes (which vary along the rows)
rowSums(prop.table(ex)) # yes & no on living standards (which vary along the columns)

```

# 2. CIs

```{r}
diffpropci.Wald.mp(b = 132, c = 107, n = 1144, conf.level = 0.95)
diffpropci.Wald.mp(b = 107, c = 132, n = 1144, conf.level = 0.95)
diffpropci.Wald.mp(b = ex[1, 2], c = ex[2, 1], n = sum(ex), conf.level = 0.95)

scoreci.mp(b = ex[1,2], c = ex[2,1], n = sum(ex), conf.level = 0.95)
```
# 3. regression

```{r}
ex_raw2 <- read.table("http://www.stat.ufl.edu/~aa/cat/data/Opinions.dat", header=TRUE)
ex_raw2

m1 <- gee(data = ex_raw2,
          formula = y ~ question,
          id = person,
          family = binomial(link = "identity"))

m1

summary(m1)

```


```{r}
m2 <- gee(data = ex_raw2,
          formula = y ~ question,
          id = person,
          family = binomial(link = "logit"))

m2

summary(m2)
```
# 4. all possible odds ratios

```{r}
df_odds <- data.frame(i = integer(101*101),
                      prop_1 = double(101*101),
                      prop_2 = double(101*101),
                      or = double(101*101))
i <- 0
for (prop_1 in seq(0, 1, 0.01)){
  for (prop_2 in seq(0, 1, 0.01)){
  i <- i + 1
  
  df_odds[i,]$i <- i
  df_odds[i,]$prop_1 <- prop_1
  df_odds[i,]$prop_2 <- prop_2
  df_odds[i,]$or <- (prop_1/(1 - prop_1))/(prop_2/(1 - prop_2))
  
  }
}

df_odds
```

```{r, fig.width=10, fig.height=10}
df_odds %>% 
  mutate(or = round(or, 1),
         or = case_when(or > 10 ~ 10,
                        or <= 10 & or >= 1 ~ or,
                        or < 1 & or >= 1 ~ 0.5,
                        or < 0.01 ~ 0)) %>% 
  ggplot(aes(x = prop_1,
             y = prop_2,
             fill = or,
             label = or)) +
  geom_raster() + 
  #geom_text(size = 1, color = "white") +
  theme_bw()

```




# --- MISC/END ---

```{r}

m_p1 <- matrix(data = rep(0:100, 101)/10,
               nrow = 101)

m_p2 <- matrix(data = rep(0:101, each = 101)/10,
                ncol = 101)

m_p1
m_p2
#m_p1 + m_p2
#m_p1 * m_p2
#round(m_p1 / m_p2, 1)

m_odds <- (m_p1 / (1 - m_p1))/(m_p2 / (1 - m_p2))

colnames(m_odds) <- c(seq(0, 1, 0.1))
rownames(m_odds) <- c(seq(0, 1, 0.1))
round(m_odds, 2)

```

